---
title: "Assignment 2"
subtitle: |
  | STAT 8670
author: |
  | Tommy Stell
  | Georgia State University
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document:
    toc: true
    number_sections: true
    df_print: kable
    highlight: tango
  rmarkdown::html_document:
    theme: readable
urlcolor: blue
fontsize: 11pt
geometry: margin=1in
---

# Question 1

Given $f(x) = \text{e}^{2x} -2$ on $x \in \mathcal{X} = [0,2]$

## 1a

Using Root-finding algorithm with $a_0 =0$ and $b_0 = 2$

Iteration 1:

First check: $f(a_0)f(b_0) = (\text{e}^{0} - 2)(\text{e}^4 -2) < 0$

$$ c_1 = {a_0+b_0 \over 2} = {0+2\over2} = 1$$

$$ f(c_1) = \text{e}^2 -2 >0 \rightarrow b_1 = c_1$$

Thus 
$$\theta_1 = (a_1=0,b_1=1,c_1=1,f(a_1)=-1,f(b_1)\approx 5.389)$$

Iteration 2:

First check: $f(a_1)f(b_1) = (\text{e}^0 - 2)(\text{e}^2-2) < 0$

$$ c_2 = {a_1+b_1 \over 2} = {0+1 \over 2} = {1\over 2}$$
$$ f(c_2) = \text{e}^1 - 2 > 0 \rightarrow b_2 = c_2$$
Thus 
$$\theta_2 = (a_2=0,b_2={1\over2},c_2={1\over2},f(a_2)=-1,f(b_2)\approx 0.718)$$

## 1b

Using NR-method with $x_0 = 0$

Given $f(x)=\text{e}^{2x}-2$, $f'(x) = 2\text{e}^{2x}$

Iteration 1:

$$x_1 = x_0 - {f(x_0) \over f'(x_0)} = 0 - {\text{e}^0 -2 \over 2 \text{e}^0} = 0 - {-1 \over 2} = {1\over 2}$$

Thus 

$$\eta_1 = (x_1 = {1\over 2}, f(x_1) \approx 0.7183)$$
Iteration 2:

$$x_2 = x_1 - {f(x_1) \over f'(x_1)} = {1\over 2} - {f\left({1\over2}\right) \over f'\left({1\over2}\right)} \approx 0.3679$$

Thus

$$\eta_2 = (x_2 \approx 0.3679, f(x_2) \approx 0.0871)$$

## 1c

Using Secant method with $x_0 =1$ and $x_{-1} = 1.5$

Iteraton 1:

$$ x_1 = x_0 - f(x_0){ x_0 - x_{-1} \over f(x_0) - f(x_{-1})} = 1-f(1){1-1.5 \over f(1)-f(1.5)} \approx 0.7878$$

Thus

$$\tau_1 = (x_1 \approx 0.7878, f(x_1)\approx 2.8336)$$

Iteration 2:

$$ x_2 = x_1 - f(x_1){x_1 - x_0 \over f(x_1) - f(x_0)} \approx 0.7878 - (2.8336){0.7878 -1 \over 2.8336 - 5.3891} \approx 0.5525$$

Thus

$$\tau_2 = (x_2 = 0.5525, f(x_2) = 1.0193 )$$


# Question 2

Using tolerance as $\delta = 10^{-5}$ and a maximum iterations of 10.


```{r}
base_function <- function(x){
    return(exp(2*x) - 2)
}

base_function_deriv <- function(x){
    return(2 * exp(2*x))
}

exact_solution <- log(2) / 2
```

## 2a 

```{r}
root_finding_method <- function(f, a, b, tol = 1e-5, max_iter = 10){
    if (f(a) * f(b) >= 0){
        stop("f(a) and f(b) must have different signs")
    }
    
    iter <- 0
    while ((b - a) / 2 > tol && iter < max_iter){
        c <- (a + b) / 2
        if (f(c) == 0){
            return(c)
        } else if (f(a) * f(c) < 0){
            b <- c
        } else {
            a <- c
        }
        iter <- iter + 1

        print(paste("Iteration:", iter)) 
        print(paste("a:", a, "b:", b, "c:", c))
        print(paste("f(a):", f(a), "f(b):", f(b)))
        cat("\n")
    }

    return((a + b) / 2)}
```

## 2b 

```{r}
NR_method <- function(f, f_deriv, x0, tol = 1e-5, max_iter = 10){
    iter <- 0
    x <- x0

    while (iter < max_iter && abs(f(x)) > tol){
        x_new <- x - f(x) / f_deriv(x)
                
        x <- x_new
        iter <- iter + 1

        print(paste("Iteration:", iter, "x:", x, "f(x):", f(x)))
    }
    
    return(x)
}
```

## 2c

```{r}
Secant_method <- function(f, x0, xn1, tol =1e-5, max_iter = 10){
    iter <- 0
    x_prev <- x0
    x_curr <- xn1

    while (iter < max_iter && abs(f(x_curr)) > tol){
        x_new <- x_curr - f(x_curr) * (x_curr - x_prev) / (f(x_curr) - f(x_prev))
        
        x_prev <- x_curr
        x_curr <- x_new
        iter <- iter + 1

        print(paste("Iteration:", iter, "x:", x_curr, "f(x):", f(x_curr)))
    }
    
    return(x_curr)
}
```

## Running the question 2 code

```{r}
# Root-finding method
print("Root-finding method:")
answer_a <- root_finding_method(base_function, 0, 2)
cat("\n")
diff_a <- abs(answer_a - exact_solution)
print(paste("Absolute Difference:", diff_a))
cat("\n")
# Newton-Raphson method
print("Newton-Raphson method:")
answer_b <- NR_method(base_function, base_function_deriv, 0)
cat("\n")
diff_b <- abs(answer_b - exact_solution)
print(paste("Absolute Difference:", diff_b))
cat("\n")
# Secant method
print("Secant method:")
answer_c <- Secant_method(base_function, 1, 1.5)
cat("\n")
diff_c <- abs(answer_c - exact_solution)
print(paste("Absolute Difference:", diff_c))

```

# Question 3

## 3a

Given $\left\{ x_n \right\}_{n \in \mathbb N}$ where $x_n=1+\left(3\over4 \right)^n$

The limit of $x_n$ is given by $\lim\limits_{n \to \infty} x_n = 1$

To check for linear convergence, there must exist a $\mu$ such that $0 \le\mu<1$ in the following:

$$ \lim\limits_{n \to \infty} {\Vert x_{n+1} - x_\infty \Vert \over \Vert x_n - x_\infty \Vert} \le \mu$$ 

In our case, 

$$\lim\limits_{n \to \infty} {\Vert x_{n+1} - x_\infty \Vert \over \Vert x_n - x_\infty \Vert} = \lim\limits_{n \to \infty} {\vert1+\left(3\over4\right)^{n+1}-1\vert \over \vert 1+\left(3\over4\right)^n-1\vert}=\lim\limits_{n\to\infty} {3 \over 4} = {3\over4}$$

So this sequence is linearly convergent. 

To check for superlinear convergence, $\mu =0$ which is not the case and thus this sequence is only linearly convergent

Thus it is not quadratic convergence either since quadratic convergence is a stronger form than superlinear convergence.

## 3b

Given $\left\{ y_n \right\}_{n \in \mathbb N}$ where $y_n=1+\left(3\over4n \right)^n$

The limit of $y_n$ is given by $\lim\limits_{n \to \infty} y_n = 1$

To check for linear convergence, there must exist a $\mu$ such that $0 \le \mu <1$ in the following:


$$ \lim\limits_{n \to \infty} {\Vert y_{n+1} - y_\infty \Vert \over \Vert y_n - y_\infty \Vert} \le \mu$$ 

In our case:

$$\lim\limits_{n \to \infty} {\Vert y_{n+1} - y_\infty \Vert \over \Vert y_n - y_\infty \Vert} = \lim\limits_{n \to \infty} {\vert1+\left(3\over4(n+1)\right)^{n+1}-1\vert \over \vert 1+\left(3\over4n\right)^n-1\vert}=\lim\limits_{n\to\infty} {3 \over 4} {n^n \over (n+1)^{(n+1)}}$$

To evaluate this limit, we can do the following:

$$ \lim\limits_{n\to \infty} {3\over 4}{n^n \over (n+1)^{(n+1)}} = \lim\limits_{n\to\infty} \left(n \over n+1\right)^n {1 \over n+1} = \lim\limits_{n\to\infty} \left(1 - {1 \over n+1}\right)^n {1 \over n+1} = \lim\limits_{n\to\infty} {e^{-1} \over n+1} = 0$$

So not only is it linearly convergent, it is in fact superlinearly convergent. 

Using a computer algebra system (Wolf Ram Alpha), it can be shown that it is not quadratically convergent. 


