---
title: "Assignment 4"
author: |
  | Tommy Stell
  | Georgia State University
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: true
    number_sections: true
    df_print: kable
    highlight: tango
  rmarkdown::html_document:
    theme: readable
  html_document:
    toc: true
    df_print: paged
subtitle: |
  | STAT 8670
urlcolor: blue
fontsize: 11pt
geometry: margin=1in
---

# Question 1

## 1a

Explain the difference between


a.) Training-test split
b.) K-fold cross-validation
c.) Leave one out cross validation


$\square$


a.) Training-test split

This is a random split of the original dataset into two disjoint datasets. This is done to allow your data set to have both something to form the model with (i.e. the training set) and further a set to score the accuracy of your trained data against (testing set). This is done so that the labeled data can be used to test the accuracy of the model without tainting the testing phase by including the data used a testing metric in the development of the model.

b.) K-fold cross-validation

This splits the dataset into $K$ equal disjoint subsets of data. It then selects a single fold (for example, the ith fold) and trains the model on the remaining folds. It then uses the ith fold left out intitially to evaulate the model. This process then cycles through so each one of the $K$ folds is used to evaulate the training on the remaining folds.

c.) Leave one out cross validation (LOOCV)

This is a variation of K-fold CV where $K=n$ so each fold contains a single data point. The model is then used to train on the other $n-1$ data points and evaluated on the data point left out.


The bias-variance trade off:

a.) Training-test split

Having a large fraction of the data sequestered into the test set gives a low variance on the trade off of a high bias. This is due to a significant decrease in the size of the training set. Conversely, it the test set is a small fraction of the total dataset, then the test error will be noisy and less meaningful.

b.) K-fold cross-validation

Low \$K\$ values lead to high bias and low variance while higher \$K\$ values lead to low bias and higher variance.

c.) Leave one out cross validation (LOOCV)

Low bias since it uses practically the entire dataset for training but variance is much higher due to the senstivity of the estimate to the single point left out.


$\blacksquare$

## 1b

Why is cross-validation often preferred over analytical MSE formulas for model-selection?

$\square$

First, this assumes that an analytical MSE formula can be derived. A lot of models I deal with may not have a specific analytical formula for the MSE. This leads to CV being preferred for cases where it may be either too labor intensive or impossible to determine and analytical formula.

Further, CV actually simulates the data for a variety of related models. This can lead to stronger models just due to the usage of an empirical estimation of the model.

$\blacksquare$

# Question 2

Consider the model

$$Y = \beta_0 + \beta_1 x + \beta_2 x^2 + \epsilon,\quad \epsilon \sim N(0, \sigma^2)$$

With $x_i \sim \text{Unif}(-2,2)$ and $\beta_0 =1, \beta_1=2, \beta_2 = -1$.

## 2a

Generate $n=60$ observations, fit polynomial models of degree $d=1,\ldots, 6$ and perform a 5-fold CV. Compute mean CV error for each $d$.

$\square$

```{r}
Q2_model <- function(n, beta0 = 1, beta1 = 2, beta2 = -1, sigma = 1) {
  X1 <- runif(n, min = -2, max = 2)
  epsilon <- rnorm(n, mean = 0, sd = sigma)
  Y <- beta0 + beta1 * X1 + beta2 * (X1^2) + epsilon
  data.frame(X1 = X1, Y = Y)
}

# question 2.a

Q2_observations <- Q2_model(60)

Q2_poly_degree <- c(1, 2, 3, 4, 5, 6)

Q2_CV <- function(data, degrees, k = 5) {
  n <- nrow(data)
  folds <- sample(rep(1:k, length.out = n))
  cv_errors <- numeric(length(degrees))
  
  for (d in seq_along(degrees)) {
    degree <- degrees[d]
    fold_errors <- numeric(k)
    
    for (fold in 1:k) {
      train_data <- data[folds != fold, ]
      test_data <- data[folds == fold, ]
      
      model <- lm(Y ~ poly(X1, degree), data = train_data)
      predictions <- predict(model, newdata = test_data)
      fold_errors[fold] <- mean((test_data$Y - predictions)^2)
    }
    
    cv_errors[d] <- mean(fold_errors)
  }
  
  names(cv_errors) <- paste0("Degree_", degrees)
  return(cv_errors)
}

print("MSE for polynomials of degree 1,...,6")
Q2_CV(Q2_observations, Q2_poly_degree)

```

$\blacksquare$

## 2b

Plot CV error vs degree $d$ and identify the optimal model.

$\square$

```{r}
Q2_plot_CV_errors <- function(cv_errors) {
  plot(as.numeric(sub("Degree_", "", names(cv_errors))), cv_errors, type = "b",
       xlab = "Polynomial Degree", ylab = "Cross-Validation MSE",
       main = "CV MSE vs Polynomial Degree")
}

Q2_plot_CV_errors(Q2_CV(Q2_observations, Q2_poly_degree))
```

From the plot, we can see that the MSE drops and approximately plateaus after degree 2 which means to avoid underfitting, we should have a degree at least 2. To avoid overfitting however, we want the minimum degree polynomial so we want the polynomial with the first degree after the "elbow". So the optimal model is of degree 2.

$\blacksquare$

## 2c

Compare your result with training MSE only (no CV). Explain the difference.

$\square$

```{r}
Q2_train_test_split <- function(data, train_frac = 0.7) {
  n <- nrow(data)
  train_indices <- sample(1:n, size = floor(train_frac * n))
  train_data <- data[train_indices, ]
  test_data <- data[-train_indices, ]
  list(train = train_data, test = test_data)
}

Q2_MSE <- function(train_data, test_data, degree) {
  model <- lm(Y ~ poly(X1, degree), data = train_data)
  predictions <- predict(model, newdata = test_data)
  mean((test_data$Y - predictions)^2)
}

Q2_main <- function() {
  data <- Q2_model(60)
  
  cv_errors <- Q2_CV(data, Q2_poly_degree)
  
  
  #Q2_plot_CV_errors(cv_errors)
  
  splits <- Q2_train_test_split(data)
  train_data <- splits$train
  test_data <- splits$test
  
  test_MSEs <- sapply(Q2_poly_degree, function(d) {
    Q2_MSE(train_data, test_data, d)
  })
  
  print("CV error of different degrees")
  names(cv_errors) <- paste0("Degree_", Q2_poly_degree)
  print(cv_errors)
  print("")
  print("MSE of different degrees")
  names(test_MSEs) <- paste0("Degree_", Q2_poly_degree)
  print(test_MSEs)
}

Q2_main()
```

Overall, the MSE is lower than the CV error of the different degrees. Both methods of calculating errors pick out a order 2-3 polynomial. Curiously, the MSE of a degree 3 model is lower than the MSE of a degree 2 model which is odd due to the true nature of the model being a degree 2 polynomial. Knowing this, the CV is more reliable because it allows us to pick a model which fits the true nature of the data better. The MSE alone fails to distinguish between the correct model and one that is overfit.

$\blacksquare$

# Question 3

For linear regression $\hat Y = HY$, where $H=X(X^TX)^{-1}X^T$.

## 3a

Show that LOOCV prediction errors can be computed without refitting

$$ \text{err}_\text{LOOCV} = {1 \over n}\sum\limits_{i=1}^n \left ( Y_i - \hat Y_{(i),i} \over 1 - h_{ii} \right)^2 $$

Where $h_{ii}$ is the $i$th leverage.

$\square$

For simplicity, let us denote the following, $X\in\mathbb{R}^{n\times p}$ is the dataset with $n$ data points and $p$ parameters, $X_{(i)}$ is the dataset without the ith observation, $H_{(i)}$ is the constructed $H$ matrix without the ith component. Further, $Y_i$ is the observed value at the ith data point, $\hat Y_i$ is the fitted value at the ith datapoint. $Y_{(i)}$ is the all the observations without $Y_i$. $Y_{(i),i}$ is the predicted value for the ith datapoint fitted without seeing the ith datapoint.

We can express $\hat Y_i$ as follows.

$$
\hat Y_i = HY =  h_{ii}Y_i + \sum\limits_{j\neq i} h_{ij} Y_j
$$ This means we can write $\hat Y_{(i),i}$ as $\hat Y_i - h_{ii}Y_i$. This is simply the fitted prediction without the ith component. We can then take the difference between the preceding terms to get the following.

$$
\hat Y_i - \hat Y_{(i),i} = h_{ii}(Y_i - \hat Y_{(i),i})
$$ $$
\implies Y_i - \hat Y_{(i),i} = {Y_i - \hat Y_i \over 1 - h_{ii}}
$$

The above can then be used to determine the MSE between the true observation at the ith datapoint and the LOO predicted value for every possible datapoint we can remove. This results in the following.

$$
\text{err}_\text{LOOCV} = {1 \over n} \sum\limits_{i=1}^n (Y_i - Y_{(i),i})^2 = {1 \over n}\sum\limits_{i=1}^n \left( {Y_i - \hat Y_i \over 1 - h_{ii}}\right)^2
$$ So we can find the LOOCV error without refitting.

$\blacksquare$

## 3b

Simulate $n=40, p=3$ features with noise $\sigma=1$. Compare LOOCV MSE computer via refitting and via the shortcut formula. Confirm equality up to rounding error.

$\square$

```{r}
Q3_n <- 40
Q3_p <- 3
Q3_sigma <- 1

Q3_X <- matrix(rnorm(Q3_n * Q3_p), nrow = Q3_n, ncol = Q3_p)
Q3_data <- data.frame(y = rnorm(Q3_n, sd = Q3_sigma), Q3_X)

Q3_fit_model <- function(data) {
  model <- lm(y ~ ., data = data)
  summary(model)
}

Q3_X_matrix <- model.matrix(lm(y ~ ., data = Q3_data))

Q3_H <- Q3_X_matrix %*% solve(t(Q3_X_matrix) %*% Q3_X_matrix) %*% t(Q3_X_matrix)

h <- diag(Q3_H)

Q3_shortcut_LOOCV <- function(data, h) {
    n <- nrow(data)
    model <- lm(y ~ ., data = data)
    residuals <- resid(model)
    loo_cv_errors <- (residuals / (1 - h))^2
    mean(loo_cv_errors)
}

Q3_refitting_LOOCV <- function(data) {
    n <- nrow(data)
    loo_cv_errors <- numeric(n)

    for (i in 1:n) {
        train_data <- data[-i, ]
        test_data <- data[i, , drop = FALSE]

        model <- lm(y ~ ., data = train_data)
        prediction <- predict(model, newdata = test_data)

        loo_cv_errors[i] <- (test_data$y - prediction)^2
    }

    mean(loo_cv_errors)
}

Q3_main <- function() {
    fit_summary <- Q3_fit_model(Q3_data)
    
    time_shortcut_start <- Sys.time()
    loocv_shortcut_mse <- Q3_shortcut_LOOCV(Q3_data, h)
    time_shortcut_end <- Sys.time()
    time_refit_start <- Sys.time()
    loocv_refit_mse <- Q3_refitting_LOOCV(Q3_data)
    time_refit_end <- Sys.time()

    absolute_difference <- abs(loocv_shortcut_mse - loocv_refit_mse)

    print(paste("LOOCV MSE (Shortcut):", loocv_shortcut_mse))
    print(paste("LOOCV MSE (Refitting):", loocv_refit_mse))
    print(paste("Absolute Difference:", absolute_difference))
    print(paste("Time taken (Shortcut):", time_shortcut_end - time_shortcut_start))
    print(paste("Time taken (Refitting):", time_refit_end - time_refit_start))
}

Q3_main()

```

We can see that the difference between the refitting and shortcut methods are the same within machine precision. The main difference is that the time for computing the MSE via the refitting was approximately two orders of magnitude larger than the time it took for the shortcut method of finding the MSE. This points towards the shortcut method being preferred for time complexity reasons.

$\blacksquare$

# Question 4

Let $X_1, \ldots, X_n \sim N(\mu,1)$ with $\mu=0$. Estimate the standard error of the sample median $\hat \theta$.

## 4a

Write the non-parametric bootstrap algorithm for estimating standard error (SE).

$\square$

Algorithm:

Step 1:

Compute $\hat \theta = \text{median}\{ X_1, \ldots, X_n \}$

Step 2:

For $1 \leq i \leq B$: Sample in $X=\{X_1, \ldots, X_n \}$, denote $X_{(i)}$ as the ith bootstrapped sample. Compute $\hat \theta_{(i)} = \text{median}(X_{(i)})$. Collect all $\hat\theta_{(i)}$

Step 3:

Compute the bootstrap standard error shown below.

$$MSE_{(bootstrap)} = \sqrt{{1 \over B-1} \sum\limits_{i=1}^B \left( \hat\theta - \hat\theta_{(i)}\right)^2}$$

$\blacksquare$

## 4b

Simulate $n=50, B=2000$. Compare bootstrap SE with theoretical asymptotic $SE \approx 1/(2f(\mu)\sqrt{n})$ where $f(\mu) = 1/\sqrt{2\pi}$.

$\square$

```{r}
Q4_sample_median_error <- function(n = 50, B = 2000) {
  medians <- numeric(B)
  
  for (b in 1:B) {
    sample_data <- rnorm(n)
    medians[b] <- median(sample_data)
  }
  
  median_estimate <- mean(medians)
  temp <- mean((medians - median_estimate)^2)

  MSE_bootstrap <- ((1/(B-1)) * sum((medians - median_estimate)^2))^0.5

  list(median_estimate = median_estimate, mse = MSE_bootstrap)

}

Q4_compute_theoretical <- function(n = 50, mu = 0, sigma = 1) {
    f_mu <- 1 / (2*pi)^0.5
    theoretical <- 1/(2*(n^0.5)*f_mu)
    return(theoretical)
}

Q4_compare_errors <- function() {
  n <- 50
  B <- 2000
  
  sim_results <- Q4_sample_median_error(n, B)
  theoretical_mse <- Q4_compute_theoretical(n)
  
  print(paste("Simulated Median Estimate:", sim_results$median_estimate))
  print(paste("Simulated MSE:", sim_results$mse))
  print(paste("Theoretical MSE:", theoretical_mse))
  print(paste("MSE error:", abs(theoretical_mse - sim_results$mse)))
}

Q4_compare_errors()
```

The bootstrap MSE is very close to the theoretical MSE. This points towards the bootstrap MSE being a good method to determining the $\hat\theta$.

$\blacksquare$

# Question 5

Let $\hat \theta = \max(X_1,\ldots,X_n)$ for $X_i \sim \text{Unif}(0,1)$.

## 5a

Derive jackknife bias and variance estimators in terms of leave one out $\hat \theta_{(i)}$.

$\square$

We have that $\hat\theta = \max(X_1, \ldots, X_n)$ for $X_i \sim \text{Unif}(0,1)$. Define the LOO estimate as follows.

$$
\hat\theta_{(i)} = \max(X_1, \ldots X_{i-1}, X_{i+1}, \ldots,X_n)
$$ And the entire jackknife estimate as:

$$
\overline{\theta_{(\cdot)}} = {1 \over n}\sum\limits_{i=1}^n \hat\theta_{(i)}
$$ Since $\hat\theta$ is a measure of the maximum of the dataset, there are two possible cases for the ordered LOO estimates.Take $X_j$ as the true maximum value of $X$. The two cases are thus either $X_j$ is the unique maximum value or there exists $X_k$ such that $X_j = X_k$.

In the case that $X_j$ is the unique maximum, then the jackknife estimator $\hat\theta_{(j)}$ is less than the estimate of all other jackknife samples. Without loss of generality, assume that this is the $k-th$ sample or that $\hat\theta_{(j)}=X_k$. For every other estimate, $\hat\theta_{(i\neq j)} = X_j$. The expected value of $\hat\theta_{(i)}$ is thus as follows.

$$
\overline{\hat\theta_{(\cdot)}} = {(n-1)X_j + X_k \over n} = X_j - {X_j - X_k \over n}
$$ The bias is thus as follows.

$$
\widehat{\text{Bias}}_{\text{jack}} = (n-1)(\overline\theta_{(\cdot)} - \hat \theta) = (n-1)\left(X_j - {X_j - X_k \over n} - X_j \right)
$$ $$
\widehat{\text{Bias}}_\text{jack} = \left(1-n \over n \right)(X_j - X_k)
$$ The jackknife bias corrected estimator is thus $\hat\theta_{\text{jack}}= \hat\theta - \widehat{\text{Bias}}_\text{jack}$. This simplfies into the following.

$$
\hat\theta_{\text{jack}}= X_j + {n-1 \over n} (X_j - X_k)
$$

The variance can be computed in the standard manner by using the following equation.

$$
\widehat{\text{Var}}_\text{jack} = {n-1 \over n}\sum\limits_{i=1}^n(\hat\theta_{(i)} - \bar\theta_{(\cdot)})^2
$$ Using similar logic as with the bias, we get the following.

$$
\widehat{\text{Var}}_\text{jack} = {n-1 \over n}\left({(n-1) \over n^2}{(X_j - X_k)^2} + \left( (n-1)(X_j-X_k) \over n\right)^2\right)
$$

$$
\widehat{\text{Var}}_\text{jack} = \left(n-1 \over n \right)^2(X_j - X_k)^2
$$

$\blacksquare$

## 5b

Simulate $n=30$ and compare jackknife bias and SE with theoretical values $\mathbb{E}[\max X]= {n \over n+1}$, $\text{Var}(\max X) = {n \over (n+2)(n+1)^2}$.

$\square$

```{r}
Q5_n <- 30

Q5_true_theta <- function(n) {
  n / (n + 1)
}

Q5_true_var <- function(n) {
  n / ((n + 1)^2 * (n + 2))
}

Q5_simulate_theta_hat <- function(n, B = 10000) {
  theta_hats <- numeric(B)
  
  for (b in 1:B) {
    sample_data <- runif(n, min = 0, max = 1)
    theta_hats[b] <- max(sample_data)
  }
  
  mean_theta_hat <- mean(theta_hats)
  var_theta_hat <- var(theta_hats)
  
  list(mean_theta_hat = mean_theta_hat, var_theta_hat = var_theta_hat)
}

Q5_main <- function() {
  n <- Q5_n
  B <- 10000
  
  sim_results <- Q5_simulate_theta_hat(n, B)
  true_theta <- Q5_true_theta(n)
  true_var <- Q5_true_var(n)
  
  theta_error = abs(true_theta - sim_results$mean_theta_hat)
  var_error = abs(true_var - sim_results$var_theta_hat)
  
  print(paste("Simulated E[theta_hat]: ", round(sim_results$mean_theta_hat, 4)))
  print(paste("True E[theta_hat]: ", round(true_theta, 4)))
  print(paste("Simulated Var(theta_hat): ", round(sim_results$var_theta_hat, 6)))
  print(paste("True Var(theta_hat): ", round(true_var, 6)))
  print("")
  print(paste("E[theta_hat] error: ", round(theta_error, 4)))
  print(paste("Var(theta_hat) error: ", round(var_error, 6)))
}

Q5_main()
```

From the above, we can see that the error is quite low, on the order of $10^{-4}$ for the expected value and $10^{-6}$ for the variance. This points towards good convergence of the jackknife bias and variance to the true values. I did not do multiple different trials to determine  

$\blacksquare$

# Question 6

## 6a

Explain the difference between the sampling distribution $F_n$, the bootstrap distribution $F_n^*$, and the population distribution $F$, and their relationship.

$\square$

So, we have that $F$ is the true population distribution which is not known. This is a fixed but unknown quantity. This represents the true distribution.

$F_n$ however, is the sampled distribution from some observed data. This is estimated by assuming the multiset of observed data of size $n$, has an equal probability of each data point being selected. This allows us to sample the distribution and build an estimate of $F$. In fact, if we allow $n \rightarrow \infty$, then $F_n \rightarrow F$. The specific form of $F_n$ is given below where $\textbf{1}$ represents the indicator function.

$$
F_n = {1 \over n} \sum\limits_{i=1}^n \textbf{1}\{X_i \leq x \}
$$

The bootstrapped distribution, $F_n^*$ is similar to the empirical distribution but it allows for us to draw samples from $F_n$. Specifically, to form the $F_n^*$, we sample a new dataset, $X^*$, from $X$ where each point can be resampled. We pull $n$ samples all iid from $F_n$ so that $X^* = \{X_1^*, \ldots, X_n^* \}$.

The connection of this allows us to say that $F\approx F_n$ and likewise, $F_n \approx F_n^*$ which implies that $F\approx F_n^*$. So $F$ is the true distribution, $F_n$ is a good approximation, and $F_n^*$ is a good and computationally feasible approximation to $F_n$.

$\blacksquare$
