---
title: "Assignment 3"
author: |
  | Tommy Stell
  | Georgia State University
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: true
    number_sections: true
    df_print: kable
    highlight: tango
  rmarkdown::html_document:
    theme: readable
  html_document:
    toc: true
    df_print: paged
subtitle: |
  | STAT 8670
urlcolor: blue
fontsize: 11pt
geometry: margin=1in
---

# Question 1

Let $\theta = \int\limits_2^4 e^{-x} dx$

## 1a

Derive $\theta$ in a closed form.

$\square$

$$\theta = \int\limits_2^4 e^{-x} dx = -e^{-x} \Big|_2^4 = (-e^{-4})-(-e^{-2}) = e^{-2} - e^{-4}$$

$\blacksquare$

## 1b

Consider the simple MC estimator with $X_i \sim \text{Unif}(2,4)$

$\hat\theta_m= (4-2) {1\over m} \sum\limits_{i=1}^m \text{e}^{-X_i}$

Derive $\mathbb{E}[\hat \theta_m], \text{Var}(\hat\theta_m)$ and a large-sample $100(1-\alpha)\%$ CI for $\theta$ using the CLT

$\square$

$$\mathbb{E}[\hat\theta_m] = 2\mathbb{E}[e^{-X}] = 2 \int\limits_2^4 e^{-x} f_{\text{U}(2,4)}dx = \int\limits_2^4 e^{-x}dx = \theta$$
$$\text{Var}(\hat\theta_m) = 4{\text{Var}(e^{-X}) \over m} = {4 \over m}\left(\mathbb{E}\left[e^{-2X}\right] - \mathbb{E}\left[e^{-X}\right]^2 \right) = {4 \over m} \left(e^{-4} - e^{-8} - \left(e^{-2} - e^{-4}\right)^2 \over 4 \right) $$

$$\text{Var}(\hat\theta_m) = {e^{-4} - e^{-8} - \left(e^{-2} - e^{-4}\right)^2 \over m} = {2(e^{-6}-e^{-8}) \over m}$$

By the large sample CLT, we have $\hat\theta_m \approx N(\theta, \text{Var}(\hat\theta_m))$. So we can approximate this Ci by the following.

$$\hat\theta_m \pm z_{1-\alpha/2} \sqrt{{2 \left(e^{-6} - e^{-8} \right)\over m}}$$

$\blacksquare$

## 1c

Simulate $m \in\{10^2, 10^3, 10^4, 10^5 \}$ and produce a log-log plot of absolute error vs $m$ with setting seeds. Fit a least-squares line to $\text{log}_{10}(\text{abs. error})$ vs $\text{log}_{10}(m)$ and report the slope.

$\square$

```{r}
theta_true <- exp(-2) - exp(-4)

m_samples <- c(1e2, 1e3, 1e4, 1e5)

theta_sample <- function(m){
    
    running_sum <- 0

    for (i in 1:m){
        set.seed(i)
        x <- runif(1, 2, 4)
        running_sum <- running_sum + exp(-x)
    }

    return((4-2)*running_sum / m)
}

m_sampled_values <- numeric(length(m_samples))

for (i in seq_along(m_samples)){
    m <- m_samples[i]
    m_sampled_values[i] <- theta_sample(m)
}

m_sampled_errors <- abs(m_sampled_values - theta_true)
results <- data.frame(m = m_samples, theta_sampled = m_sampled_values, error = m_sampled_errors)
print(results)

plot(log10(results$m), log10(results$error), type='b', xlab='log10(m)', ylab='log10(error)', main='Log-Log plot of Error vs m')
abline(lm(log10(results$error) ~ log10(results$m)), col='red')

slope <- coef(lm(log10(results$error) ~ log10(results$m)))[2]
cat("Slope of the line in log-log plot:", slope, "\n")

```

$\blacksquare$

# Question 2

Let $X \sim \text{Exp}(1)$ and $Y=\text{e}^{-X}$ so that $\theta = \mathbb{E}[Y]=\mathbb{E}[e^{-X}]= \int_0^\infty e^{-x}e^{-x}dx=1/2$. Use $h(X) = X$ as a control variate (with $\mathbb{E}[h]=1, \text{Var}(h)=1$). 

## 2a 

Compute $\beta^*$, the optimal control coefficient.

$\square$

$$\beta^* = {\text{Cov}(Y,h) \over \text{Var}(h)} = \text{Cov}(e^{-X},X) = \mathbb{E}[Xe^{-X}] - \mathbb{E}[X]\mathbb{E}[e^{-X}]$$

$$\mathbb{E}[Xe^{-X}] = \int\limits_0^\infty xe^{-x}e^{-x} dx = \int\limits_0^\infty xe^{-2x}dx = {1\over 4}$$

Thus, $\beta^*$ is given by the following.

$$\beta^* = \mathbb{E}[Xe^{-X}] - \mathbb{E}[X]\mathbb{E}[e^{-X}] = {1\over 4} - {1\over 2} = - {1\over 4}$$

$\blacksquare$

## 2b 

Compute $\text{Var}(Y)$, the minimized variance with control variate, and the relative efficiency.

$\square$

For the $\text{Var}(Y)$, we can do the following.

$$\text{Var}(Y) = \mathbb{E}[Y^2] - \mathbb{E}[Y]^2 = \mathbb{E}[e^{-2X}] - \mathbb{E}[e^{-x}]^2= \int\limits_0^\infty e^{-3x}dx - \left(1\over 2\right)^2 = {1\over 3} - {1\over 4} = {1\over 12}$$

The minimized variance with control variate is given by the following.

$$\text{Var}_{min} = \text{Var}(Y) - {\text{Cov} (Y,h)^2 \over \text{Var}(h)}$$

Thus, using the previously calculated values, we get the following.

$$\text{Var}_{min} = {1\over 12} - \left( - {1\over 4} \right)^2 = {1\over 12} - {1\over 16} = {1 \over 48}$$

The relative efficiency is given by the ration of the variance to the minimized variance $\text{Var}(Y) / \text{Var}_{min}$. Thus,

$$\text{Relative Efficiency} = {\text{Var}(Y) \over \text{Var}_{min}} = {1/12 \over 1/48} = 4$$

$\blacksquare$

# Question 3

Let $\theta = \int_0^1 e^x dx = e-1$ and $g(u)=e^u$.

## 3a

Derive $\mathbb{E}[g(U)]$ and $\text{Var}(g(U))$ for $U \sim \text{Unif}(0,1)$.

$\square$

$$\mathbb{E}[g(U)] = \int\limits_0^1 e^u du = e^1 - e^0 = e-1$$

$$\mathbb{E}[g(U)^2] = \int\limits_0^1 e^{2u} du = {e^2 - 1 \over 2}$$

$$\text{Var}(g(U)) = \mathbb{E}[g(U)^2] - (\mathbb{E}[g(U)])^2 = {e^2 - 1 \over 2} - (e-1)^2 \approx 0.242$$

$\blacksquare$

## 3b

Show that $g(U)$ and $g(1-U)$ are negatively correlated and conclude that the antithetic estimator has a smaller variance than the plain MC.

$\square$

$$\mathbb{E}[g(U)g(1-U)] = \mathbb{E}[e^Ue^{1-U}]= \mathbb{E}[e^1]=e$$

$$\text{Cov}(g(U),g(1-U)) = \mathbb{E}[g(U)g(1-U)] - \mathbb{E}[g(U)] \mathbb{E}[g(1-U)]$$

Now, $\mathbb{E}[g(1-U)] = \int_0^1 e^{1-u}du = e-1$, so we can write the covariance as follows.

$$\text{Cov}(g(U),g(1-U)) = e - (e-1)^2 \approx - 0.234$$

To compute the variance of the antithetic estimator, we can do the following.

$$\text{Var}\left(g(U)+g(1-U) \over 2 \right) = {\text{Var}(g(U)) + \text{Var}(g(1-U)) + 2 \text{Cov}(g(U),g(1-U)) \over 4} $$

$$\text{Var}\left(g(U)+g(1-U) \over 2 \right) = {{e^2-1} - 2(e-1)^2 + 2e-2(e-1)^2 \over 4} = {e^2 - 1 + 2e - 4(e-1)^2 \over 4} \approx 0.0039 $$

In comparison, we can compute the variance of the average of two variables with the same distribution.

$$\text{Var}\left(g(U)+g(U) \over 2 \right) = {\text{Var}(g(U)) + \text{Var}(g(U)) + 2 \text{Cov}(g(U),g(U)) \over 4} =  { e^2-1 - 2(e-1)^2\over 4} \approx 0.121$$

So we can conclude that the variance is lower when using the antithetic variables.

$\blacksquare$

## 3c

With $m=1000$ simulate both estimators $R=100$ times and report the sample variances. Comment on the findings.

$\square$

```{r}
function_3 <- function(x){
    return(exp(-x))
}

true_3 <- exp(0) - exp(-1)

m_3 <- 1e3
R_3 <- 1e2

# simple MC

simple_MC_3 <- function(m){
    running_sum <- 0

    for (i in 1:m){
        set.seed(i)
        x <- runif(1, 0, 1)
        running_sum <- running_sum + function_3(x)
    }

    return(running_sum / m)
}

simple_MC_values <- simple_MC_3(m_3)

# antithetic MC


antithetic_MC_3 <- function(m){
    running_sum <- 0

    for (i in 1:(m/2)){
        set.seed(i)
        u <- runif(1, 0, 1)
        running_sum <- running_sum + (function_3(1-u) + function_3(u))/2
    }

    return(running_sum / m)
}

antithetic_MC_values <- antithetic_MC_3(m_3)


simple_MC_variances <- numeric(R_3)
antithetic_MC_variances <- numeric(R_3)

for (r in 1:R_3) {
    simple_MC_variances[r] <- var(replicate(100, simple_MC_3(m_3)))
    antithetic_MC_variances[r] <- var(replicate(100, antithetic_MC_3(m_3)))
}
cat("Simple MC estimate:", simple_MC_values, "with variance:", mean(simple_MC_variances), "\n")
cat("Antithetic MC estimate:", antithetic_MC_values, "with variance:", mean(antithetic_MC_variances), "\n")

```

As evident from the sample variances, we can see that the antithetic MC has a much lower variance by orders of magnitude. This points towards the usage of an anthithetic MC estimate to be much stronger.

$\blacksquare$

# Question 4

Let $U \sim \text{Unif}(0,1)$ and define $Y_1 = g(U)$ and $Y_2=g(1-U)$ for a measurable $g:[0,1] \rightarrow \mathbb{R}$. Consider the antithetic estimator

$$\hat\theta_\text{AV} = {1 \over 2m} \sum\limits_{i=1}^m \left(g(U_i) + g(1-U_i) \right), \quad U_i \stackrel{\text{i.i.d.}}{\sim} \text{Unif}(0,1$$

## 4a

Show that $\hat \theta_\text{AV}$ is unbiased for $\theta = \mathbb{E}[g(U)]$

$\square$

An estimator is unbiased iff $\mathbb{E}[\hat\theta - \theta]=0$. We can show this by the following.

$$\mathbb{E}[\hat\theta_m] = {1\over 2m}\sum\limits_{i=1}^m  \mathbb{E}[g(U_i)] + \mathbb{E}[g(1-U_i)] = {\mathbb{E}[g(U)] + \mathbb{E}[g(1-U)] \over 2 }$$

But, we know that since $U \sim \text{Unif}(0,1)$ then $\mathbb{E}[g(U)] = \mathbb{E}[g(1-U)] = \theta$.So we can simplify to the following.

$$\mathbb{E}[\hat\theta_m] ={\mathbb{E}[g(U)] + \mathbb{E}[g(1-U)] \over 2 } = {\theta + \theta \over 2} = \theta $$

Thus the bias given by $\mathbb{E}[\hat\theta_m] - \mathbb{E}[\theta]$ is exactly 0.

$\blacksquare$

## 4b

Let $\rho = \text{Corr}(g(U), g(1-U))$. Show that

$$\text{Var}(\hat\theta_\text{AV}) = {1 \over 2m} \text{Var}(g(U))(1+\rho)$$

When does antithetic sampling reduce variance versus plain MC using 2m i.i.d. draws?

$\square$

We have the following.

$$\text{Var}(\hat\theta_\text{AV}) = \text{Var}\left({1 \over 2m} \sum\limits_{i=1}^m \left(g(U_i) + g(1-U_i) \right) \right) = {\text{Var}(g(U) +g(1-U)) \over m}$$

Examining the inner variance, we get the following.

$$\text{Var}(g(U) +g(1-U)) = {  \text{Var}(g(U)) + \text{Var}(g(1-U)) + 2 \text{Cov}(g(U),g(1-U))  \over 4}$$

But we know that $\text{Var}(g(U)) = \text{Var}(g(1-U))$. This allows us to rewrite as follows.

$$\text{Var}(g(U) +g(1-U)) = {  2\text{Var}(g(U))  + 2 \rho \text{Var}(g(U))  \over 4} = {\text{Var}(g(U))\over 2}(1+\rho)$$

Thus, we can say that $\text{Var}(\hat\theta_\text{AV})$ is given by the following.

$$\text{Var}(\hat\theta_\text{AV}) ={\text{Var}(g(U))\over 2m}(1+\rho)$$
From the above, this will only reduce the variance if $\rho < 0$ which will lead to a decrease in the variance.

$\blacksquare$

# Question 5

## 5a

Suppose $g$ is monotone on $[0,1]$. Determine the sign of $\rho=\text{Corr}(g(U),g(1-U))$ for (i) monotone increasing $g$, (ii) monotone decreasing $g$. Give an intuitive explanation.

$\square$

We know that the correlation is directly related to the covariance so to investigate the bahvior of $\rho$, it is sufficient to study the behavior of the covariance.

In other words, we need to show that the covariance is always negative regardless if $g$ is monotonically increasing or decreasing. Without loss of generality, assume $g$ is monotonically increasing. This can be said by the following.

$$ x<y \implies f(x)\leq f(y)\quad \text{or} \quad f(y) - f(x) \geq 0 $$

This in turns leads to the following statement.

$$x<y \iff 1-x > 1-y \implies f(1-y) - f(1-x) \leq 0$$
This means the product of the two preceding statements l(eads to the following.

$$\left(f(y) - f(x) \right)\left( f(1-y) - f(1-x) \right) \leq 0\quad \forall x,y \in [0,1]$$
Take the double integral of this expression across the unit square.

$$I=\int\limits_0^1 \int\limits_0^1 (f(y) - f(x))(f(1-y)-f(1-x))dx dy$$

$$I = \int\limits_0^1 \int\limits_0^1 f(y)f(1-y) - f(y)f(1-x)-f(x)f(1-y) + f(x)f(1-x) dx dy$$
$$I = \int\limits_0^1 \int\limits_0^1 f(y)f(1-y)dx dy - \int\limits_0^1 \int\limits_0^1f(y)f(1-x) dx dy-\int\limits_0^1 \int\limits_0^1f(x)f(1-y)dx dy + \int\limits_0^1 \int\limits_0^1f(x)f(1-x) dx dy$$
This can be simplified significantly by recognizing that the functions are symmetric and thus we can split this into the following sum, $I = I_1 + I_2 + I_3 + I_ 4$, and then analyze each part individually. 

$$I_1 = \int\limits_0^1 \int\limits_0^1 f(y)f(1-y)dx dy = \int\limits_0^1 f(y)f(1-y)dy$$
$$I_4 = \int\limits_0^1 \int\limits_0^1 f(x)f(1-x)dx dy = \int\limits_0^1 f(x)f(1-x)dx$$
So we can say that by the proper change of variables, $I_1 = I_4$. We can write this form of the integral as follows.

$$I_a = \int\limits_0^1 f(x)f(1-x)dx$$


Now taking $I_2$ and $I_3$, we see the following since the interiors are independent, we can seperate the integrals.

$$I_2 = \int\limits_0^1 \int\limits_0^1f(y)f(1-x) dx dy = \left( \int\limits_0^1 f(y) dy \right)\left( \int\limits_0^1 f(1-x) dx \right)$$
$$I_3 = \int\limits_0^1 \int\limits_0^1f(x)f(1-y) dx dy = \left( \int\limits_0^1 f(x) dx \right)\left( \int\limits_0^1 f(1-y) dy \right)$$

Which, by the proper change of variables again, we get the that $I_2 = I_3$.We can write this form of the integral as follows.

$$I_b = \left( \int\limits_0^1 f(x) dx \right)\left( \int\limits_0^1 f(1-x) dx \right)$$

Thus, we can write $I=2I_a - 2I_b$, or less succinctly,

$${I \over 2} = \int\limits_0^1 f(x)f(1-x)dx - \left( \int\limits_0^1 f(x) dx \right)\left( \int\limits_0^1 f(1-x) dx \right)$$
Which the term $I/2$ is the covariance given by $\text{Cov}(g(u),g(1-u))$. Thus we can say that regardless if the function $g$ is monotonically increasing or monotonically decreasing, we get that the covariance is always negative.


$\blacksquare$
